Lab 4: Graphical interpretation of LMs: Answer Key

### 2023-01-13 Lab 4  Graphical Interpretation of LMs/Lab4_Graphical_interpretation_LM.Rmd
Add files via upload

---
title: "Lab 4: Graphical interpretation of LMs"
author: "Eric Parajon"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include = TRUE)

@@

#Part 1: Ggplot visualizations
Lets begin by returning to the mtcars data set and creating a basic scatterplot of displacement (disp) vs mpg.

@@

Okay great, but what if we wanted to plot the linear relationship between our x and y?

@@

Right now, we're using color as the aesthetic mapped to a specific explanatory variable (color represents the number of cylinders). We can play around with this:

@@

What are the benefits/drawbacks of each of these strategies? Think about how well the plot communicates information about the types of variables plotted as well as the relationship between them.

The above are great places to start out visualizing any potential relationships between your variables.

Now let's jump back into the palmerpenguins dataset to perform some more in-depth visualization of regression models.

Here we will evaluate the association between penguin body mass (measured in grams) and flipper length (measured in mm).

Let's suppose our main research question is to determine whether body mass was associated with flipper length We will set the body mass as our dependent variable and flipper length as our independent variable (or predictor of interest).

<div align="left">$\large E[Body_mass_{i} | Flipper_length_{i}] = \beta_0 + \beta_1  Flipper_length_{i} + \epsilon,$</div>

where $E[Body_mass_{i} |  Flipper_length_{i}]$ denotes the expected body mass for penguin $i$ given the flipper length of penguin $i$.

@@

Now, we can construct a linear regression model with body mass as the dependent variable and flipper length. as the independent variable (or predictor of interest).

@@

```
Lets again plot it but this time break out the results by species.

@@

where $E[Body\_mass_{i} |  Flipper\_length_{i}], Species{i}]$ denotes the expected body mass for penguin $i$ given the flipper length of penguin $i$ and controlling for species of the penguin $i$.

@@

Okay great, but what if we want to visualize substantive effects? For this we can use our old friend predict()!

@@

subeffects_plot <- ggplot(int_dat, aes(x=flipper_length_mm, y=fit, ymin=lwr, ymax=upr,color=species)) +
        geom_pointrange() +
 scale_color_manual(values = c("darkorange","purple","cyan4")) +
  labs(title = "Flipper length and body mass",
       x = "Flipper length (mm)",
       y = "Body mass (g)",
       color = "Penguin species",
       shape = "Penguin species")
subeffects_plot+
  plot_theme

@@

## Interaction analysis
Additionally, in other models we may be interested in exploring the interaction between variables. Often these interaction effects are difficult to interpret without graphic aids.

In this example, suppose we are interested in knowing if the effect of body mass on bill length is attenuated by the sex of the penguin. We begin by plotting the data and then perform interaction analysis.

Our regression equation looks like this:

@@

```{r}
#First let's plot the raw data to see if there's anything there.
penguins %>% 
  # drop rows with missing data
  drop_na(sex) %>% 
  ggplot(aes(x = body_mass_g, y = bill_length_mm, color = sex)) +
  geom_point() +
  facet_wrap(~ sex)

@@

#Part 3: LM practice
Use the msleep dataset to model how body weight (bodywt) and brain weight (brainwt) influences total REM sleep (sleep_rem). Then create a model including an additional predictor (of your choice). Create a coefficient plot of the results of the two models and compare them.

@@

#Interpret 
```

>>>>

---
title: 'Lab 4: Graphical interpretation of LMs'
author: "Eric Parajon"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include = TRUE, echo=TRUE, eval=FALSE)

@@

# Part 1: Ggplot visualizations

Lets begin by returning to the mtcars data set and creating a basic scatterplot of displacement (disp) vs mpg.

@@

+ Okay great, but what if we wanted to plot the linear relationship between our x and y?

@@

+ Right now, we're using color as the aesthetic mapped to a specific explanatory variable (color represents the number of cylinders). We can play around with this:

@@

+ What are the benefits/drawbacks of each of these strategies? Think about how well the plot communicates information about the types of variables plotted as well as the relationship between them.

+ The above are great places to start out visualizing any potential relationships between your variables.

2. Now let's jump back into the palmerpenguins dataset to perform some more in-depth visualization of regression models.

+ Here we will evaluate the association between penguin body mass (measured in grams) and flipper length (measured in mm).

+ Let's suppose our main research question is to determine whether body mass was associated with flipper length We will set the body mass as our dependent variable and flipper length as our independent variable (or predictor of interest):

<div align="left">$\large E[Body\_mass_{i} | Flipper\_length_{i}] = \beta_0 + \beta_1  Flipper\_length_{i} + \epsilon,$</div>

+ where $E[Body\_mass_{i} |  Flipper\_length_{i}]$ denotes the expected body mass for penguin $i$ given the flipper length of penguin $i$.

@@

+ Now, we can construct a linear regression model with body mass as the dependent variable and flipper length. as the independent variable (or predictor of interest).

@@

```
+ Lets again plot it but this time break out the results by species.

@@

+ where $E[Body\_mass_{i} |  Flipper\_length_{i}], Species{i}]$ denotes the expected body mass for penguin $i$ given the flipper length of penguin $i$ and controlling for species of the penguin $i$.

@@

+ Okay great, but what if we want to visualize substantive effects? For this we can use our old friend predict()!

@@

subeffects_plot <- ggplot(int_dat, aes(x=flipper_length_mm, y=fit, ymin=lwr, ymax=upr,color=species)) +
        geom_pointrange() +
 scale_color_manual(values = c("darkorange","purple","cyan4")) +
  labs(title = "Flipper length and body mass",
       x = "Flipper length (mm)",
       y = "Body mass (g)",
       color = "Penguin species",
       shape = "Penguin species")+
  plot_theme

@@

## Interaction analysis
+ Additionally, in other models we may be interested in exploring the interaction between variables. Often these interaction effects are difficult to interpret without graphic aids.

+ In this example, suppose we are interested in knowing if the effect of body mass on bill length is attenuated by the sex of the penguin. We begin by plotting the data and then perform interaction analysis.

+ Our regression equation looks like this:

@@

```{r}
#First let's plot the raw data to see if there's anything there.
penguins %>%  
  drop_na(sex) %>%  # drop rows with missing data
  ggplot(aes(x = body_mass_g, y = bill_length_mm, color = sex)) +
  geom_point() +
  facet_wrap(~ sex)

@@

# Part 3: **Exercise** LM practice

1.  Use the msleep dataset to model how body weight (bodywt) and brain weight (brainwt) influences total REM sleep (sleep_rem). 
2.  Then create a model including an additional predictor (of your choice). 
3. Create a coefficient plot of the results of the two models and compare them.

@@

#Interpret the graph
```

### 2023-01-13 Lab 4  Graphical Interpretation of LMs/Lab4_Graphical_interpretation_LM_Answers.Rmd
Add files via upload

---
title: "Lab 4: Graphical interpretation of LMs"
author: "Eric Parajon"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include = TRUE)

@@

#Part 1: Ggplot visualizations
Lets begin by returning to the mtcars data set and creating a basic plot of Displacement vs mpg.

@@

Okay great, but what if we wanted to plot the linear relationship between our x and y?

```{r}
carsplot + geom_smooth(method = "lm", se = FALSE, 
                       color = "darkgrey", size = 0.75)
#geom_smooth is for fitted lines.
#se=TRUE plots confidence intervals
#We can adjust the color and size of the line plotted.
#Also, note that we just added a layer over carsplot - did we change the carsplot object itself?
```

Right now, we're using color as the aesthetic mapped to a specific explanatory variable (color represents the number of cylinders). We can play around with this:

@@

#Okay great so huge effect! But what about confounders? Perhaps the penguins species is also important.

```
Lets again plot it but this time break out the results by species.

@@

Okay great, but what if we want to visualize substantive effects? For this we can use our old friend predict()!

@@

subeffects_plot <- ggplot(int_dat, aes(x=flipper_length_mm, y=fit, ymin=lwr, ymax=upr,color=species)) +
        geom_pointrange() +
 scale_color_manual(values = c("darkorange","purple","cyan4")) +
  labs(title = "Flipper length and body mass",
       x = "Flipper length (mm)",
       y = "Body mass (g)",
       color = "Penguin species",
       shape = "Penguin species")
subeffects_plot+
  plot_theme

## Interaction analysis
Additionally, in other models we may be interested in exploring the interaction between variables. Often these interaction effects are difficult to interpret without graphic aids.

In this example, suppose we are interested in knowing if the effect of body mass on bill length is attenuated by the sex of the penguin. We begin by plotting the data and then perform interaction analysis.

Our regression equation looks like this:

@@

```
#Part 3: LM practice
Use the msleep dataset to model how body weight (bodywt) and brain weight (brainwt) influences total REM sleep (sleep_rem). Then create a model including an additional predictor (of your choice). Create a coefficient plot of the results of the two models and compare them.

@@

#Plotting (make sure to assign color and shape to the model variable in the ggplot call)
sleep_results_plot <- ggplot(data=sleep_results, aes(x=term, y=estimate, 
           color = model, shape = model)) + 
    geom_hline(yintercept=0, color="red", size=.5) +
    geom_errorbar(aes(ymin=conf.low, ymax=conf.high, width=0), size=.5, position=position_dodge(width=0.5)) +
    geom_point(aes(y=estimate), size=1.75, position = position_dodge(width=0.5)) +
  labs(title = "Model Estimates of Brain and Body Weight (and total sleep) on REM Sleep",
       x = "Predictor",
       y = "Coefficient Estimate",
       caption = "Models fit with OLS. Error bars show the 95% confidence interval.",
       color = "Model:",
       shape = "Model:") +
   scale_x_discrete(labels = c("Body Weight", "Brain Weight", "Total Sleep"))+ 
  coord_flip()+
  plot_theme

>>>>

---
title: 'Lab 4: Graphical interpretation of LMs: Answer Key'
author: "Eric Parajon"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include = TRUE, echo=TRUE, eval=TRUE)

@@

# Part 1: Ggplot visualizations
Lets begin by returning to the mtcars data set and creating a basic plot of Displacement vs mpg.

@@

+ Okay great, but what if we wanted to plot the linear relationship between our x and y?

```{r}
carsplot + geom_smooth(method = "lm", se = FALSE, 
                       color = "darkgrey", linewidth = 0.75)
#geom_smooth is for fitted lines.
#se=TRUE plots confidence intervals
#We can adjust the color and size of the line plotted.
#Also, note that we just added a layer over carsplot - did we change the carsplot object itself?
```

+ Right now, we're using color as the aesthetic mapped to a specific explanatory variable (color represents the number of cylinders). We can play around with this:

@@

Penguin_results
#Okay great so huge effect! But what about confounders? Perhaps the penguins species is also important.

```
+ Lets again plot it but this time break out the results by species.

@@

Penguin_results_2
```

+ Okay great, but what if we want to visualize substantive effects? For this we can use our old friend predict()!

@@

subeffects_plot <- ggplot(int_dat, aes(x=flipper_length_mm, y=fit, ymin=lwr, ymax=upr,color=species)) +
        geom_pointrange() +
 scale_color_manual(values = c("darkorange","purple","cyan4")) +
  labs(title = "Flipper length and body mass",
       x = "Flipper length (mm)",
       y = "Body mass (g)",
       color = "Penguin species",
       shape = "Penguin species")+
  plot_theme

subeffects_plot
```

## Interaction analysis
+ Additionally, in other models we may be interested in exploring the interaction between variables. Often these interaction effects are difficult to interpret without graphic aids.

+ In this example, suppose we are interested in knowing if the effect of body mass on bill length is attenuated by the sex of the penguin. We begin by plotting the data and then perform interaction analysis.

+ Our regression equation looks like this:

@@

```
# Part 3: **Exercise** LM practice

1.  Use the msleep dataset to model how body weight (bodywt) and brain weight (brainwt) influences total REM sleep (sleep_rem). 
2.  Then create a model including an additional predictor (of your choice). 
3. Create a coefficient plot of the results of the two models and compare them.

@@

#Plotting (make sure to assign color and shape to the model variable in the ggplot call)
sleep_results_plot <- ggplot(data=sleep_results, aes(x=term, y=estimate, 
           color = model, shape = model)) + 
    geom_hline(yintercept=0, color="red", size=.5) +
    geom_errorbar(aes(ymin=conf.low, ymax=conf.high, width=0), linewidth=.5, position=position_dodge(width=0.5)) +
    geom_point(aes(y=estimate), size=1.75, position = position_dodge(width=0.5)) +
  labs(title = "Model Estimates of Brain and Body Weight (and total sleep) on REM Sleep",
       x = "Predictor",
       y = "Coefficient Estimate",
       caption = "Models fit with OLS. Error bars show the 95% confidence interval.",
       color = "Model:",
       shape = "Model:") +
   scale_x_discrete(labels = c("Body Weight", "Brain Weight", "Total Sleep"))+ 
  coord_flip()+
  plot_theme

sleep_results_plot
```
